---
title: "IST Trial Data Processing"
author: "Mike Krähenbühl"
date: "2025-06-29"
output: html_document
---


# Introduction

The code for data processing is taken from (https://gitlab.uzh.ch/hongruyu.chen/causal_ml-for-ite). The aim is to use the same pre-processed data and apply our models to evaluate wheter we come to the same conclusions as Chen et al. (2025).

The individual patient data was downloaded from (https://trialsjournal.biomedcentral.com/articles/10.1186/1745-6215-12-101#Sec8) and saved in the file IST_dataset_2011.csv. In this script this raw file is processed to create the dataset for the ITE estimation.


# Set Up

```{r Setup}
library("here")
library("tidyverse")
library("broom")
library("progress")
library("CalibrationCurves")
library("gtools")
```



# Load Raw Data

Variables names and comments can be found at <https://trialsjournal.biomedcentral.com/articles/10.1186/1745-6215-12-101/tables/2>.

```{r load data}

# load csv data as data frame
IST.dataset <- read_csv(here("data", "IST_dataset_2011.csv")) %>% 
  as_tibble()
```

# Data Overview

```{r data overview}
str(IST.dataset)
head(IST.dataset)
dim(IST.dataset)
```


# Quality Check

```{r check data consistency and quality}
# check aspirin given within 14 days
IST.dataset %>% select(RXASP, DASP14) %>% filter(RXASP != DASP14) %>% filter(RXASP == "N" & DASP14 == "Y")

# check time on trial treatment
IST.dataset %>% select(RXASP, ONDRUG) %>% summary()


# check baseline characteristics
summary(IST.dataset$RDELAY)
summary(IST.dataset$AGE)
summary(IST.dataset$RSBP)
table(IST.dataset$SEX, useNA = "always")
table(IST.dataset$RCT, useNA = "always")
table(IST.dataset$RVISINF, useNA = "always")
table(IST.dataset$RATRIAL, useNA = "always")
table(IST.dataset$RASP3, useNA = "always")
table(IST.dataset$RDEF1, useNA = "always")
table(IST.dataset$RDEF2, useNA = "always")
table(IST.dataset$RDEF3, useNA = "always")
table(IST.dataset$RDEF4, useNA = "always")
table(IST.dataset$RDEF5, useNA = "always")
table(IST.dataset$RDEF6, useNA = "always")
table(IST.dataset$RDEF7, useNA = "always")
table(IST.dataset$RDEF8, useNA = "always")
table(IST.dataset$RCONSC, useNA = "always")
table(IST.dataset$STYPE, useNA = "always")
table(IST.dataset$RXASP, useNA = "always")

# check non-trial treatments
IST.dataset %>% select(DAP, DOAC, DTHROMB, DGORM, DSTER, DCAA) %>% mutate_if(is.character, as.factor) %>% summary() # most patients don't have these treatments

# check final diagnosis at 14 days
IST.dataset %>% select(DDIAGISC, DDIAGHA, DDIAGUN, DNOSTRK) %>% mutate_if(is.character, as.factor) %>% summary() 

# check death within 14 days
IST.dataset %>% select(DDEAD) %>% mutate_if(is.character, as.factor) %>% summary() 
table(IST.dataset$DDEAD, IST.dataset$DPLACE, useNA = "always")

# check restroke within 14 days
table(IST.dataset$DRSISC, useNA = "always")
table(IST.dataset$DRSH, useNA = "always")
table(IST.dataset$DRSUNK, useNA = "always")
IST.dataset %>% select(DRSISC, DRSH, DRSUNK) %>% 
  mutate(DRESTROKE = if_else(DRSISC == "Y" | DRSH == "Y" | DRSUNK == "Y", 1, NA)) %>%
  mutate(DRESTROKE = if_else(DRSISC == "N" & DRSH == "N" & DRSUNK == "N", 0, DRESTROKE)) %>%
  mutate(DRESTROKE = as.factor(DRESTROKE)) %>% select(DRESTROKE) %>% summary()
table(IST.dataset$STRK14, useNA = "always") # estimated by IST group

# check outcome at 6 months
# check death/dependence at 6 months
IST.dataset %>% select(FDEAD, FDENNIS) %>% mutate_if(is.character, as.factor) %>% summary() 

# use %in% can pass NA to next if_else condition check; use == can't pass NA to next condition check and will directly return NA
test <- IST.dataset %>% 
  mutate(FDEAD = if_else(FDEAD %in% "U", NA, FDEAD),
         FDENNIS = if_else(FDENNIS %in% "U", NA, FDENNIS)) %>% 
  mutate(OUTCOME6M = if_else(FDEAD %in% "Y" | FDENNIS %in% "Y", 1, if_else((FDEAD %in% "N" | is.na(FDEAD)) & is.na(FDENNIS), NA, 0))) %>% 
  select(DDEAD,FDEAD, FDENNIS, OCCODE, FRECOVER, OUTCOME6M) %>% 
  mutate(across(, as.factor))

summary(test)

(test %>% 
  filter(OCCODE %in% c(1,2) & (OUTCOME6M == 0 | is.na(OUTCOME6M)) ))

(test %>% 
  filter(FDEAD %in% c("N"), is.na(FDENNIS)))

(test %>% 
  filter(FRECOVER == "Y" & OUTCOME6M == 1))

(test %>% filter(FDEAD=="Y", FDENNIS == "N")) # can keep it since the outcome is correct

(test %>% filter(DDEAD=="Y", FDEAD=="N")) # can keep it since the outcome is correct

(test %>% filter(FDEAD=="Y"))
```

# Data Complete

```{r create region and outcomes variable}
# create region variable as a factor based on country
IST.dataset.new <- IST.dataset %>% 
  mutate(REGION = if_else(CNTRYNUM %in% c(43,2,3,4,7,8,9,10,11,12,31,36,14,39,15,17,18,19,33,44,20,22,24,25,27), "Europe", 
                         if_else(CNTRYNUM %in% c(29,42,6), "South America",
                                if_else(CNTRYNUM %in% c(5,28), "North America",
                                       if_else(CNTRYNUM %in% c(37,23), "South Asia",
                                              if_else(CNTRYNUM %in% c(30,41,38,40,34,26), "East Asia",
                                                     if_else(CNTRYNUM %in% c(32,13,35), "Middle East",
                                                            if_else(CNTRYNUM %in% c(1,16), "Oceania",
                                                                   if_else(CNTRYNUM %in% c(21), "Africa", "Other"))))))))) %>%
  mutate(REGION = as.factor(REGION)) %>% 
  # death within 14 days
  mutate(DDEAD = if_else(DDEAD %in% "U", NA, DDEAD)) %>% 
  # restroke within 14 days
  mutate(DRESTROKE = if_else(DRSISC == "Y" | DRSH == "Y" | DRSUNK == "Y", 1, NA)) %>%
  mutate(DRESTROKE = if_else(DRSISC == "N" & DRSH == "N" & DRSUNK == "N", 0, DRESTROKE)) %>% 
  # death / dependence at 6 months
  mutate(FDEAD = if_else(FDEAD %in% "U", NA, FDEAD),
         FDENNIS = if_else(FDENNIS %in% "U", NA, FDENNIS)) %>% 
  mutate(OUTCOME6M = if_else(FDEAD %in% "Y" | FDENNIS %in% "Y", 1, if_else((FDEAD %in% "N" | is.na(FDEAD)) & is.na(FDENNIS), NA, 0))) 
  

summary(IST.dataset.new$REGION)
summary(IST.dataset.new$OUTCOME6M)
```

```{r create dataset with 6 months outcome}
IST.dataset.clean.6m <- IST.dataset.new %>% 
  select(RXASP, AGE, RDELAY, RSBP, SEX, RCT, RVISINF, RATRIAL, RASP3, 
         RDEF1, RDEF2, RDEF3, RDEF4, RDEF5, RDEF6, RDEF7, RDEF8, RCONSC, 
         STYPE, REGION, OUTCOME6M) 

factor_cols = c("RXASP", "SEX", "RCT", "RVISINF", "RATRIAL", "RASP3", 
                "RDEF1", "RDEF2", "RDEF3", "RDEF4", "RDEF5", "RDEF6", "RDEF7", 
                "RDEF8", "RCONSC", "STYPE", "REGION")

IST.dataset.clean.6m <- IST.dataset.clean.6m %>% 
  mutate(across(all_of(factor_cols), factor))

head(IST.dataset.clean.6m)
```

```{r create dataset with 14 days outcome - death}
IST.dataset.clean.14d <- IST.dataset.new %>% 
  select(RXASP, AGE, RDELAY, RSBP, SEX, RCT, RVISINF, RATRIAL, RASP3, 
         RDEF1, RDEF2, RDEF3, RDEF4, RDEF5, RDEF6, RDEF7, RDEF8, RCONSC, 
         STYPE, REGION, DDEAD) %>% 
  mutate(DDEAD = ifelse(DDEAD %in% "Y", 1 , ifelse(DDEAD %in% "N", 0, DDEAD))) %>% 
  mutate(DDEAD= as.numeric(DDEAD)) %>%
  rename(OUTCOME14D = DDEAD)

factor_cols = c("RXASP", "SEX", "RCT", "RVISINF", "RATRIAL", "RASP3", 
                "RDEF1", "RDEF2", "RDEF3", "RDEF4", "RDEF5", "RDEF6", "RDEF7", 
                "RDEF8", "RCONSC", "STYPE", "REGION")

IST.dataset.clean.14d <- IST.dataset.clean.14d %>% 
  mutate(across(all_of(factor_cols), factor))

head(IST.dataset.clean.14d)
```


```{r create dataset with 14 days outcome - death or restroke}
IST.dataset.clean.14d.dr <- IST.dataset.new %>% 
  select(RXASP, AGE, RDELAY, RSBP, SEX, RCT, RVISINF, RATRIAL, RASP3, 
         RDEF1, RDEF2, RDEF3, RDEF4, RDEF5, RDEF6, RDEF7, RDEF8, RCONSC, 
         STYPE, REGION, DDEAD, DRESTROKE) %>% 
  mutate(DDEAD = ifelse(DDEAD %in% "Y", 1 , ifelse(DDEAD %in% "N", 0, DDEAD))) %>% 
  mutate(OUTCOME14D = if_else(DDEAD == 1 | DRESTROKE == 1, 1, NA)) %>% 
  mutate(OUTCOME14D = if_else(DDEAD == 0 & DRESTROKE == 0, 0, OUTCOME14D)) %>%
  mutate(OUTCOME14D = as.numeric(OUTCOME14D)) %>%
  select(-DDEAD, -DRESTROKE)

factor_cols = c("RXASP", "SEX", "RCT", "RVISINF", "RATRIAL", "RASP3", 
                "RDEF1", "RDEF2", "RDEF3", "RDEF4", "RDEF5", "RDEF6", "RDEF7", 
                "RDEF8", "RCONSC", "STYPE", "REGION")

IST.dataset.clean.14d.dr <- IST.dataset.clean.14d.dr %>% 
  mutate(across(all_of(factor_cols), factor))

head(IST.dataset.clean.14d.dr)
```

# Save the Processed Data

```{r save data}
# save data as R data
save(IST.dataset.clean.6m, IST.dataset.clean.14d, IST.dataset.clean.14d.dr, file = here("data","IST_model_data.RData"))
```



# End

```{r record session info}
sessionInfo()
```
